Revealing Transient Strain in Geodetic Data with Gaussian Process Regression

### 2017-06-20 HinesHetlandGPR.tex
added more

Most methods for estimating crustal strain rates from GNSS data assume some parametric form of the deformation signal. The simplest method for estimating secular crustal strain rates assumes that GNSS derived velocities can be described with a first order polynomial (i.e., having constant deformation gradients) over some subnetwork of the GNSS stations \citep[e.g.,][]{Feigl1990,Murray2000}. The components of the strain rate tensor  for each subnetwork are then determined through a least squares fit to the observations. The assumption that deformation gradients are spatially uniform is not appropriate when subnetworks span a large area.  To help overcome this deficiency, \citet{Shen1996,Shen2015} used an inverse distance weighting scheme, in which the estimated strain rate at some point is primarily controlled by observations at nearby stations. However, the methods described in \citet{Shen1996,Shen2015} are still formulated by assuming that the deformation gradients are uniform over the entire network. The errors in this assumption manifest as implausibly low formal uncertainties for the estimated strain rates. Other methods for estimating secular strain rates have parameterized GNSS derived velocities with bi-cubic splines \citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and elastostatic Green's functions \citep{Sandwell2016}. The type of basis functions and the number of degrees of freedom for a parameterization can be subjective. If there are too few degrees of freedom in the parameterization, then estimated strain rates will be biased and the uncertainties will be underestimated. On the other hand, if there are too many degrees of freedom, then there will not be any coherent features in the estimated strain rates. The methods described by \citet{Beavan2001}, \citet{Tape2009}, and \citet{Sandwell2016} also require the user to specify penalty parameters that control a similar trade-off between bias and variance in the solution. One could parameterize deformation with a physically motivated model of interseismic deformation \citep[e.g.,][]{Meade2005,McCaffrey2007}. In such models the lithospheric rheology and fault geometries are assumed to be known. Any errors in the assumed physical model could similarly result in biased strain estimates and underestimated formal uncertainties.

@@

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient crustal strain rates. Our attention is limited to horizontal strain rates in this study. We denote transient crustal strain rates as $\dot{\du{\varepsilon}}(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We determine $\dot{\du{\varepsilon}}(p)$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$.  As we discuss in Section \ref{sec:NoiseModel} and also as argued for in XXX, estimates of $\dot{\du{\varepsilon}}$ are more effective at illuminating geophysical signal than estimates of $\vec{u}(p)$. We make a prior assumption that each component of $\vec{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(p) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(p,p')$ is a covariance function indicating how we expect $u_i(p)$ to covary with $u_i(p')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We drop the component subscripts with the understanding that the same analysis is being repeated to estimate the easting and northing components of $\vec{u}$. We assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((\vec{x},t),(\vec{x}',t')\right) = X(\vec{x},\vec{x}')T(t,t').
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in Section \ref{sec:SignalModel}.

@@

Another significant source of noise in GNSS data is common mode error \citep[e.g.,][]{Wdowinski1997,Dong2006}, which is noise that is highly spatially correlated. When not accounted for, common mode error manifests as spatially uniform undulations in $\hat{u}$. However, we are primarily interested in estimating strain rates, which are insensitive to common mode error. We therefore do not include common mode error in our noise model. We then make the simplifying assumption that $\eta(p) = 0$ for the easting and northing component of GNSS data.

@@

Next we identify which covariance function for $T$ best describes the SSEs. One approach is to compare the REML likelihood for each covariance function, similar to the analysis in \citet{Langbein2004}. In Table 1, we summarize how the log REML likelihoods for the Wendland and IBM covariance functions compare to the SE covariance function.  Based on the differences in log REML likelihoods, the data is substantially more likely to come from the SE or Wendland covariance functions than from the IBM covariance function. The REML likelihoods do not definitively indicate whether the SE or Wendland covariance function is preferable.

To further explore which covariance function for $T$ best describes SSEs, we compare the observations to the predicted displacements for each covariance function. We consider the data prediction vector to be $\hat{\bm{d}} = \left(u(\bm{P}) + \bm{G}\bm{a}\right)|\bm{d}_*$. It can be shown that $\hat{\bm{d}}$ is normally distributed with mean 
\begin{equation}\label{eq:DataPredMean}
\bm{\mu}_{\hat{d}} = \left[\begin{array}{cc}
                           C_u(\bm{P},\bm{P}) & \bm{G} \\
                           \end{array}\right]
                     \left[\begin{array}{cc}
                           \bm{\Sigma} & \bm{G} \\
                           \bm{G}^T  & \bm{0} \\
                           \end{array}\right]^{-1}
                     \left[\begin{array}{c}
                           \bm{d}_* \\
                           \bm{0} \\
                           \end{array}\right]
\end{equation}  
and covariance
\begin{equation}\label{eq:DataPredCov}
\bm{C}_{\hat{\bm{d}}} = C_u(\bm{P},\bm{P}) - 
                        \left[\begin{array}{cc}
                              C_u(\bm{P},\bm{P}) & \bm{G} \\
                              \end{array}\right]
                        \left[\begin{array}{cc}
                              \bm{\Sigma} & \bm{G} \\
                              \bm{G}^T  & \bm{0} \\
                              \end{array}\right]^{-1}
                        \left[\begin{array}{c}
                              C_u(\bm{P},\bm{P}) \\
                              \bm{G}^T \\
                              \end{array}\right].
\end{equation}
We compute $\hat{\bm{d}}$ using SE, Wendland, and IBM covariance functions for $T$ and the median hyperparameters from Table 1. Figure \ref{fig:Fit} compares the easting component of $\bm{d}_*$ to $\hat{\bm{d}}$ for the Winter 2016 SSE at Station P436. The data prediction vector appears to accurately describe displacements throughout the SSE, regardless of the choice of $T$. Since the SSE signal is strongest at Station P436, it can be assumed that $T$ adequately describes the signal elsewhere. The prediction for the IBM covariance function contains slightly more high frequency, and perhaps spurious, features. The predictions for the Wendland and SE covariance functions are nearly indistinguishable. In our estimates of transient strain discussed in the next section, we ultimately settle on the Wendland covariance function for $T$ and use the median values from Table 1 for the hyperparameters. We choose the Wendland covariance function over the SE covariance function because of its computational advantages.

@@

\subsection{Transient Strain Rates}\label{sec:Results} 
Having established a noise model and a prior for transient displacements, we use the cleaned GNSS dataset to calculate transient strain rates in the Puget Sound region.  We calculate transient strain rates for each day from January 1, 2010 to March 15, 2017. The strain rates are estimates at a grid of points spanning the study area. In Figure \ref{fig:StrainMap} we show the transient strain rates on January 1, 2016, which coincides with the height of an SSE. We have included a video showing the map view of strain rates over time as supplementary material. The strain rates shown in Figure \ref{fig:StrainMap} are generally similar to the strain rates for the other six SSEs considered in this study. The SSEs cause compression in the Olympic Peninsula and extension east of Puget Sound. For comparison, estimated secular strain rates indicate trench perpendicular compression throughout this study region \citep{Murray2000,McCaffrey2007,McCaffrey2013}. The SSE are thus concentrating tectonically accumulated strain energy trench-ward, and presumably pushing the subduction zone closer to failure. Similar conclusions have been drawn based on fault slip models \citep[e.g.,][]{Dragert2001,Wech2009,Schmidt2010}, which reveal that SSEs are occurring down-dip of the seismogenic zone. A key difference between the strain inferred here and strain derived from fault slip models is that our estimated strain rates are not based on an assumed physical model. In contrast, fault slip models can be biased by errors in the assumed fault geometry or lithospheric rheology. Moreover, the degrees of freedom in fault slip models usually cannot be constrained by GNSS data alone, and it is necessary to impose regularization which further biases the results. Since our estimated strain rates lack such systematic errors, we can be more confident that our solution is unbiased and has meaningful uncertainties.

\begin{figure*}
\includegraphics{figures/strain_map/strain-map.pdf}
\caption{Estimated transient strain rates during the Winter 2016 SSE. Strain glyphs show the normal strain rate along each azimuth, where orange indicates compression and blue indicates extension. The shaded regions indicate one standard deviation uncertainties in the normal strain rate.}   
\label{fig:StrainMap}
\end{figure*}

@@

\begin{figure*}
\includegraphics{figures/strain_ts/mag-ts.pdf}
\caption{(Top) Signal-to-noise ratio (eq. \ref{eq:SNR}) at the position shown in Figure \ref{fig:Context}. (bottom) Frequency of tremors in the region shown in Figure \ref{fig:Context}.}   
\label{fig:StrainMag}
\end{figure*}

@@

There is potential for a more thorough analysis of the spatio-temporal noise in GNSS data, $\eta$, than what was performed in Section \ref{sec:NoiseModel}. In this study, we did not explored the spatial covariance of $\eta$. Spatially correlated common mode noise \citep{Wdowinski1997}, which results from factors such as reference frame error, is a non-negligible component of GNSS data. We ignore common mode error in this study because transients strain rates are insensitive to it. For other geophysical studies based on GNSS data, such as fault slip inversions, it may be necessary to incorporate a spatially covarying noise model \citep[e.g.,][]{Miyazaki2003}. We can also improve upon the seasonal model used in this study, which consists of four spatially uncorrelated sinusoids for each station. We did not explore the spatial covariance of seasonal deformation or the temporal ``roughness'' (i.e., the number of sinusoids needed to describe the observations). The periodic Gaussian process \citep{Mackay1998} is an alternative model for seasonal deformation and is well suited for exploring the roughness of seasonal deformation.  The periodic Gaussian process has zero mean and covariance function,
\begin{equation}\label{eq:Periodic}
T(t,t') = \phi \exp\left(\frac{-\sin(\pi|t - t'|)^2}{2\tau^2}\right).
\end{equation}
Realizations have annual periodicity and the roughness is controlled by $\tau$. Decreasing $\tau$ has the same effect as including higher frequency sinusoids in the seasonal model. The optimal value for $\tau$ can be found with the REML method as described in Section \ref{sec:NoiseModel}.

We have estimated transient strain rates constrained by about seven years of daily displacement observations from 94 GNSS stations. It can be computationally intensive to evaluate eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) for a dataset with this size. We significantly reduced the amount of memory needed to estimate transient strain rates by describing the temporal covariance of displacements with a compact Wendland covariance function. By using a compact covariance function for our prior, the matrix being inverted in eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) becomes sparse. We then evaluate eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) with CHOLMOD, which is designed for solving sparse positive definite systems of equations. The matrix being inverted in eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) is not positive definite; however, we can use another partitioned matrix inversion identity from \citet[e.g.,][]{Press2007} to partition it into positive definite submatrices. Even when using a compact covariance function, it may still be necessary to reduce the computational burden by dividing the data into subsets and evaluate transient strain rates for each subset.

\section{Conclusion}\label{sec:Conclusion}
In this paper we have discussed how Gaussian process regression (GPR) can be used to estimate transient strain rates from GNSS data. Most other methods for estimating strain from GNSS data assume a parametric representation of deformation, which can bias the estimated strain rates if the parameterization is not chosen carefully. The method described in this paper assumes a stochastic prior model for displacements, which can be chosen objectively with the Restricted Maximum Likelihood method. Because GPR is a Bayesian method, the uncertainties on our estimated transient strain rates are well quantified, allowing one to discern geophysical signal from noise. This is demonstrated in our application to Cascadia, where we detect transient strain resulting from slow slip events. GPR is not robust against outliers, which are common in GNSS data, and so we also introduced an effective pre-processing method for identifying and removing outliers from GNSS datasets. Another complication with GPR is that it usually involves inverting a dense covariance matrix with rank equal to the number of observations. This can be prohibitive when using years of daily GNSS observations from a network of several hundred stations.  We overcome this complication by using compact Wendland covariance function for our prior model, which allows us to utilize sparse matrix software. We believe that Gaussian process regression is a powerful tool that could be applied to a wider range of geophysical problems than what was considered in this paper.

% In the previous section we demonstrate that Gaussian processes can be an effective tool for illuminating 
% GPR can be used for more than just identifying transient strain rates and

% discuss potential applications for slip models

% discuss how we remove errors like reference frame and seasonals

% discuss how this can be used for tikhonov regularization

% we do not interpolate the strain field because it may introduce spurious artifacts Baxter2011

@@

%The material presented in this paper is primarily focused on GPS data; however, we speculate that the RBF-FD scheme can be of particular use in denoising borehole strain meter (BSM) data.  The Plate Boundary Observatory has deployed X BSMs along the Western United States.  BSM data contains low frequency drift resulting from relaxation of the borehole \citep{Gladwin1987}, which can obscure the geophysical signal of interest.  We suggest that the RBF-FD scheme may be useful in denoising BSM data.  Since the RBF-FD scheme provides a straight-forward mapping from GPS displacements to strain at any target locations, it is possible to use GPS derived strains as \textit{a priori} information for strain at BSM sites. GPS derived strains could then aid in discerning tectonic signal from noise in BSM data.

%Additional uses for strain: transient detection, prior for strain meters,

%Additional potential applications of the RBF-FD method: regularizing inverse problems,

% cite eric calais for triangulation strain

>>>>

Most methods for estimating crustal strain rates from GNSS data assume some parametric form of the deformation signal. The simplest method for estimating secular crustal strain rates assumes that GNSS derived velocities can be described with a first order polynomial (i.e., having constant deformation gradients) over some subnetwork of the GNSS stations \citep[e.g.,][]{Feigl1990,Murray2000}. The components of the strain rate tensor  for each subnetwork are then determined through a least squares fit to the observations. The assumption that deformation gradients are spatially uniform is not appropriate when subnetworks span too large of an area. To help overcome this deficiency, \citet{Shen1996,Shen2015} used an inverse distance weighting scheme, in which the estimated strain rate at some point is primarily controlled by observations at nearby stations. However, the methods described in \citet{Shen1996,Shen2015} are still formulated by assuming that the deformation gradients are uniform over the entire network. The errors in this assumption manifest as implausibly low formal uncertainties for the estimated strain rates. Other methods for estimating secular strain rates have parameterized GNSS derived velocities with bi-cubic splines \citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and elastostatic Green's functions \citep{Sandwell2016}. The type of basis functions and the number of degrees of freedom for a parameterization can be subjective. If there are too few degrees of freedom in the parameterization, then estimated strain rates will be biased and the uncertainties will be underestimated. On the other hand, if there are too many degrees of freedom, then there will not be any coherent features in the estimated strain rates. The methods described by \citet{Beavan2001}, \citet{Tape2009}, and \citet{Sandwell2016} also require the user to specify penalty parameters that control a similar trade-off between bias and variance in the solution. One could parameterize deformation with a physically motivated model of interseismic deformation \citep[e.g.,][]{Meade2005,McCaffrey2007}. In such models the lithospheric rheology and fault geometries are assumed to be known. Any errors in the assumed physical model could similarly result in biased strain estimates and underestimated formal uncertainties.

@@

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient crustal strain rates. Our attention is limited to horizontal strain rates in this study. We denote transient crustal strain rates as $\dot{\du{\varepsilon}}(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We determine $\dot{\du{\varepsilon}}(p)$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. We make a prior assumption that each component of $\vec{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(p) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(p,p')$ is a covariance function indicating how we expect $u_i(p)$ to covary with $u_i(p')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We drop the component subscripts with the understanding that the same analysis is being repeated to estimate the easting and northing components of $\vec{u}$. We assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((\vec{x},t),(\vec{x}',t')\right) = X(\vec{x},\vec{x}')T(t,t').
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in Section \ref{sec:SignalModel}.

@@

Another significant source of noise in GNSS data is common mode error \citep[e.g.,][]{Wdowinski1997,Dong2006}, which is noise that is highly spatially correlated. When not accounted for, common mode error manifests as spatially uniform undulations in estimated transient displacements. However, estimated transient strain rates are insensitive to common mode error. We therefore do not include common mode error in our noise model. We then make the simplifying assumption that $\eta(p) = 0$ for the easting and northing component of GNSS data.

@@

Next we identify which covariance function for $T$ best describes the SSEs. One approach is to compare the REML likelihood for each covariance function, similar to the analysis in \citet{Langbein2004}. In Table 1, we summarize how the log REML likelihoods for the Wendland and IBM covariance functions compare to the SE covariance function.  Based on the differences in log REML likelihoods, the data is substantially more likely to come from a Gaussian process with a SE or Wendland covariance function than an IBM covariance function. The REML likelihoods do not definitively indicate whether the SE or Wendland covariance function is preferable.

To further explore which covariance function for $T$ best describes SSEs, we compare the observations to the predicted displacements for each covariance function. We consider the data prediction vector to be $\hat{\bm{d}} = \left(u(\bm{P}) + \bm{G}\bm{a}\right)|\bm{d}_*$. It can be shown that $\hat{\bm{d}}$ is normally distributed with mean 
\begin{equation}\label{eq:DataPredMean}
\bm{\mu}_{\hat{d}} = \left[\begin{array}{cc}
                           C_u(\bm{P},\bm{P}) & \bm{G} \\
                           \end{array}\right]
                     \left[\begin{array}{cc}
                           \bm{\Sigma} & \bm{G} \\
                           \bm{G}^T  & \bm{0} \\
                           \end{array}\right]^{-1}
                     \left[\begin{array}{c}
                           \bm{d}_* \\
                           \bm{0} \\
                           \end{array}\right]
\end{equation}  
and covariance
\begin{equation}\label{eq:DataPredCov}
\bm{C}_{\hat{\bm{d}}} = C_u(\bm{P},\bm{P}) - 
                        \left[\begin{array}{cc}
                              C_u(\bm{P},\bm{P}) & \bm{G} \\
                              \end{array}\right]
                        \left[\begin{array}{cc}
                              \bm{\Sigma} & \bm{G} \\
                              \bm{G}^T  & \bm{0} \\
                              \end{array}\right]^{-1}
                        \left[\begin{array}{c}
                              C_u(\bm{P},\bm{P}) \\
                              \bm{G}^T \\
                              \end{array}\right].
\end{equation}
We compute $\hat{\bm{d}}$ using SE, Wendland, and IBM covariance functions for $T$ and the median hyperparameters from Table 1. Figure \ref{fig:Fit} compares the easting component of $\bm{d}_*$ to $\hat{\bm{d}}$ for the Winter 2015-2016 SSE at Station P436. The data prediction vector appears to accurately describe displacements throughout the SSE, regardless of the choice of $T$. Since the SSE signal is strongest at Station P436, it can be assumed that $T$ adequately describes the signal elsewhere. The prediction for the IBM covariance function contains slightly more high frequency, and perhaps spurious, features. The predictions for the Wendland and SE covariance functions are nearly indistinguishable. In our estimates of transient strain discussed in the next section, we ultimately settle on the Wendland covariance function for $T$ and use the median values from Table 1 for the hyperparameters. We choose the Wendland covariance function over the SE covariance function because of its computational advantages.

@@

\subsection{Transient Strain Rates}\label{sec:Results} 
Having established a noise model and a prior for transient displacements, we use the cleaned GNSS dataset to calculate transient strain rates in the Puget Sound region.  We calculate transient strain rates for each day from January 1, 2010 to March 15, 2017. The strain rates are estimates at a grid of points spanning the study area. In Figure \ref{fig:StrainMap} we show the transient strain rates on January 1, 2016, which coincides with the height of an SSE. We have included a video showing the map view of strain rates over time as supplementary material. The strain rates shown in Figure \ref{fig:StrainMap} are generally similar to the strain rates for the other six SSEs considered in this study. The SSEs cause compression in the Olympic Peninsula and extension east of Puget Sound. For comparison, estimated secular strain rates indicate trench perpendicular compression throughout this study region \citep{Murray2000,McCaffrey2007,McCaffrey2013}. The SSEs are thus concentrating tectonically accumulated strain energy trench-ward, and presumably pushing the subduction zone closer to failure. Similar conclusions have been drawn based on fault slip models \citep[e.g.,][]{Dragert2001,Wech2009,Schmidt2010}, which reveal that SSEs are occurring down-dip of the seismogenic zone and migrating stress upward. A key difference between the strain inferred here and strain that can be derived from fault slip models is that our estimated strain rates are not based on an assumed physical model. In contrast, fault slip models can be biased by errors in the assumed fault geometry or lithospheric rheology. Moreover, the degrees of freedom in fault slip models usually cannot be constrained by GNSS data alone, and it is necessary to impose regularization which further biases the results. Since our estimated strain rates lack such systematic errors, we can be more confident that our solution is unbiased and has meaningful uncertainties.

\begin{figure*}
\includegraphics{figures/strain_map/strain-map.pdf}
\caption{Estimated transient strain rates during the Winter 2015-2016 SSE. Strain glyphs show the normal strain rate along each azimuth, where orange indicates compression and blue indicates extension. The shaded regions indicate one standard deviation uncertainties in the normal strain rates.}   
\label{fig:StrainMap}
\end{figure*}

@@

\begin{figure*}
\includegraphics{figures/strain_ts/mag-ts.pdf}
\caption{(top) Signal-to-noise ratio (eq. \ref{eq:SNR}) at the position shown in Figure \ref{fig:Context}. (bottom) Frequency of tremors in the region shown in Figure \ref{fig:Context}.}   
\label{fig:StrainMag}
\end{figure*}

@@

There is potential for a more thorough analysis of the spatio-temporal noise in GNSS data, $\eta$, than what was performed in Section \ref{sec:NoiseModel}. In this study, we did not explored the spatial covariance of $\eta$. Spatially correlated common mode noise, which results from factors such as reference frame error, is a non-negligible component of GNSS data. We ignore common mode error in this study because transients strain rates are insensitive to it. For other geophysical studies based on GNSS data, such as fault slip inversions, it may be necessary to incorporate a spatially covarying noise model \citep[e.g.,][]{Miyazaki2003}. We can also improve upon the seasonal model used in this study, which consists of four spatially uncorrelated sinusoids for each station. We did not explore the spatial covariance of seasonal deformation or the temporal ``roughness'' (i.e., the number of sinusoids needed to describe the observations). The periodic Gaussian process \citep{Mackay1998} is an alternative model for seasonal deformation and is well suited for exploring the roughness of seasonal deformation.  The periodic Gaussian process has zero mean and the covariance function
\begin{equation}\label{eq:Periodic}
T(t,t') = \phi \exp\left(\frac{-\sin(\pi|t - t'|)^2}{2\tau^2}\right).
\end{equation}
Realizations have annual periodicity and the roughness is controlled by $\tau$. Decreasing $\tau$ has the same effect as including higher frequency sinusoids in the seasonal model. The optimal value for $\tau$ can be found with the REML method as described in Section \ref{sec:NoiseModel}.

The transient strain rates estimated in this study are constrained by about seven years of daily displacement observations from 94 GNSS stations. It can be computationally intensive to evaluate eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) for a dataset with this size. We significantly reduce the amount of memory needed to estimate transient strain rates by describing the temporal covariance of displacements with a compact Wendland covariance function. Using a compact covariance function for our prior turns eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) into sparse systems of equations, which we then solve with CHOLMOD. CHOLMOD is designed for solving sparse, positive definite systems of equations. The matrix being inverted in eqs. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) is not positive definite; however, we can use another partitioned matrix inversion identity from \citet{Press2007} to partition it into positive definite submatrices to be inverted. Even when using a compact covariance function, it may still be necessary to reduce the computational burden by dividing the data into subsets and evaluating transient strain rates for each subset.

\section{Conclusion}\label{sec:Conclusion}
In this paper we propose using Gaussian process regression (GPR) to estimate transient strain rates from GNSS data. Most other methods for estimating strain rates assume a parametric representation of deformation, which can bias the results if the parameterization is not chosen carefully. Here we assume a stochastic, rather than parametric, prior model for displacements. Our prior model describes how much we expect transient displacements to covary spatially and temporally. If we know nothing about the underlying signal that we are trying to recover, then the prior model can be chosen objectively with maximum likelihood methods. Because GPR is a Bayesian method, the uncertainties on our estimated transient strain rates are well quantified, allowing one to discern geophysical signal from noise. We demonstrate that GPR is an effective tool for detecting geophysical phenomena, such as slow slip events, in our application to GNSS data from Cascadia. One limitation with GPR is that it is not robust against outliers. To overcome this limitation, we have introduced an effective pre-processing method for identifying and removing outliers from GNSS datasets. Another complication with GPR is that it usually involves inverting a dense matrix where the number of rows and columns is equal to the number of observations. This is prohibitive when using several years of daily GNSS observations from a network of several hundred stations. We significantly reduce the computational burden of GPR by using compact Wendland covariance function to describe our prior model. While this paper just focuses on estimating transient strain rates, we believe that GPR is a powerful tool that can be applied to a wide range of geophysical problems.

\section{Acknowledgements}
This material is based upon work supported by the National Science Foundation under grant EAR 1245263. This material is based on EarthScope Plate Boundary Observatory data services provided by UNAVCO through the GAGE Facility with support from the National Science Foundation (NSF) and National Aeronautics and Space Administration (NASA) under NSF Cooperative Agreement EAR-1261833.

@@
